How to Not write Virtual Machines for Dynamic Languages

### 2007-05-27 talk/dyla2007/dyla.tex
the proper magic to be able to use utf-8

\documentclass{llncs}

\usepackage{makeidx}
\usepackage{graphicx}

@@

Dynamic languages are traditionally implemented by writing a virtual
machine for them, centered around an interpreter and/or a built-in
compiler and providing the object model and memory management. When
a language becomes more successful, the limitations of such an implementation
lead to the emergence of alternative implementations that try to solve some
of the problems. Another reason for new implementations is the desire to have
the language integrate well with existing, well-tuned object-oriented virtual
machine like Java's. In this paper, we describe the mechanisms that lead to an
abundance of implementations, and explore some of the limitations of standard
VMs.  We propose a different complementary alternative: flexibly generating
virtual machines instead
of writing them by hand.  This approach has been validated by the PyPy
project, with which we can automatically insert many features into the
generated virtual machines – including good just-in-time compilers tuned
to the dynamic language at hand.
\footnote{This research is partially supported by the EU funded %research
 project:
IST 004779 PyPy (PyPy: Implementing Python in Python).} \\

@@

For these reasons writing a virtual machine in C has many problems because it
forces the language implementer to deal with many low-level details. Limitations
of the C implementation lead to alternative implementations which draw
work-power from the reference implementation. An alternative to writing
implementations in C is to build them on top of one of the newer object oriented
virtual machines (``OO VM'') such as the JVM or the CLR. This is often wanted by
the community anyway, since it leads to better re-usability of libraries of
these platforms. However, if a C implementation existed before the
implementation of such a VM is started, this enters in conflict with the goal of
having to maintain essentially a single, simple enough implementation for a
given programming language: as the language becomes popular, there will be a
demand for having it run on various platforms – high-level VMs as well as
C-level environments.

The argument we will make in the present paper is that it is possible to
benefit from and integrate with OO VMs while keeping the dynamic
language implemented by a single, simple source code base.  The idea is
to write an interpreter for that language in another sufficiently
high-level but less dynamic language.  This interpreter plays the role
of a specification for the dynamic language.  With a good enough
translation toolchain we can then generate whole virtual machines from
this specification – either full custom VMs for C-level operating
systems, or layers on top of various OO VMs.  In other words,
meta-programming techniques can be used to successfully replace a
foreseeable one-VM-fits-all standardization attempt.

>>>>

\documentclass[utf8x]{llncs}

\usepackage{makeidx}
\usepackage{graphicx}
\usepackage[utf8x]{inputenc}

@@

Dynamic languages are traditionally implemented by writing a virtual
machine for them, centered around an interpreter and/or a built-in
compiler and providing the object model and memory management. When
a language becomes more successful, the limitations of such an implementation
lead to the emergence of alternative implementations that try to solve some
of the problems. Another reason for new implementations is the desire to have
the language integrate well with existing, well-tuned object-oriented virtual
machine like Java's. In this paper, we describe the mechanisms that lead to an
abundance of implementations, and explore some of the limitations of standard
VMs.  We propose a different complementary alternative: flexibly generating
virtual machines instead
of writing them by hand.  This approach has been validated by the PyPy
project, with which we can automatically insert many features into the
generated virtual machines -- including good just-in-time compilers tuned
to the dynamic language at hand.
\footnote{This research is partially supported by the EU funded %research
 project:
IST 004779 PyPy (PyPy: Implementing Python in Python).} \\

@@

For these reasons writing a virtual machine in C is problematic because it
forces the language implementer to deal with many low-level details. Limitations
of the C implementation lead to alternative implementations which draw
work-power from the reference implementation. An alternative to writing
implementations in C is to build them on top of one of the newer object oriented
virtual machines (``OO VM'') such as the JVM or the CLR. This is often wanted by
the community anyway, since it leads to better re-usability of libraries of
these platforms. However, if a C implementation existed before the
implementation of such a VM is started, this enters in conflict with the goal of
having to maintain essentially a single, simple enough implementation for a
given programming language: as the language becomes popular, there will be a
demand for having it run on various platforms -- high-level VMs as well as
C-level environments.

The argument we will make in the present paper is that it is possible to
benefit from and integrate with OO VMs while keeping the dynamic
language implemented by a single, simple source code base.  The idea is
to write an interpreter for that language in another sufficiently
high-level but less dynamic language.  This interpreter plays the role
of a specification for the dynamic language.  With a good enough
translation toolchain we can then generate whole virtual machines from
this specification -- either full custom VMs for C-level operating
systems, or layers on top of various OO VMs.  In other words,
meta-programming techniques can be used to successfully replace a
foreseeable one-VM-fits-all standardization attempt.
